\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex
Tweaks

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale, in real time.
Building on the open-source \texttt{tskit} library, the method
explicitly reconstructs genealogies with both
mutation and recombination, which
can be conveniently and efficiently analysed using standard Python data science
tools. As illustrated in Figure~\ref{fig:overview_sc2ts}, inference is based on incrementally adding batches of sequences
based on their collection dates and proceeds in three phases.
First, possible paths connecting each sample to the current ARG
are inferred (allowing for recombination) using the
Li and Stephens (LS) model (Figure~\ref{fig:overview_sc2ts}A, B);
the LS ``copying process'' is a Hidden Markov Model (HMM)
approximating the effects of mutation
and recombination, widely used in large-scale genomics (Section~\ref{sec:ls}).
Then, since many samples
in a batch can share an attachment path,
we infer phylogenetic trees for each of these clusters separately using standard methods
(Figure~\ref{fig:overview_sc2ts}C; Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample clusters to the current
ARG and apply some parsimony-based heuristics to address issues
introduced by the inherent greediness of this strategy
(Figure~\ref{fig:overview_sc2ts}D, E; Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts}, we infer two
large ARGs (with 1,265,685 and 657,239 samples, respectively) and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development.
We find that these
ARGs accurately capture known phylogenetic
relationships (Section~\ref{sec:backbone_phylogeny}) and
mutational spectra (Section~\ref{sec:mutation_spectrum}),
as well as automatically identifying the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages})
with a high level of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of accurate joint estimation of
genetic inheritance with mutation and recombination events will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

@@

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans that recombine in every generation,
recombination is relatively rare in SARS-CoV-2, with recombination nodes accounting
for $<$0.3\% of the inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of strains. We expect ARGs to be particularly tree-like
early in the pandemic, when co-infection was less likely and divergence between
lineages relatively low.

@@

The \texttt{sc2ts} backbone phylogenies for these Nextstrain subsamples contain small amounts of recombination,
with 7 recombination nodes in the Wide ARG backbone (8 for the Long ARG
backbone). However, the backbones are highly tree-like, as these
recombination events involve minor, local topological rearrangements
where recombination only occurs between close relatives.
% sometimes even within a single lineage such that the parent
% lineages above a recombination event coalesce with each other before involvement with
% any other part of the phylogeny.
These local reticulations could either represent
real but previously unobserved recombination events between closely related strains
or could be artefacts of the current \texttt{sc2ts} method and its parameterization.

>>>>

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale, in real time.
Building on the open-source \texttt{tskit} library, the method
explicitly reconstructs genealogies with both
mutation and recombination, which
can be conveniently and efficiently analysed using standard Python data science
tools. As illustrated in Figure~\ref{fig:overview_sc2ts}, inference is based on incrementally adding batches of sequences
based on their collection dates and proceeds in three phases.
First, possible paths connecting each sample to the current ARG
are inferred (allowing for recombination) using the
Li and Stephens (LS) model (Figure~\ref{fig:overview_sc2ts}A, B);
the LS ``copying process'' is a Hidden Markov Model (HMM)
approximating the effects of mutation
and recombination, widely used in large-scale genomics (Section~\ref{sec:ls}).
Then, since many samples
in a batch can share an attachment path,
we infer phylogenetic trees for each of these clusters separately using standard methods
(Figure~\ref{fig:overview_sc2ts}C; Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample clusters to the current
ARG and apply some parsimony-based heuristics to address issues
introduced by the inherent greediness of this strategy
(Figure~\ref{fig:overview_sc2ts}D, E; Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts}, we infer two
large ARGs (with 1,265,685 and 657,239 samples, respectively) and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development.
We find that these
ARGs accurately capture known phylogenetic
relationships (Section~\ref{sec:backbone_phylogeny}) and
mutational spectra (Section~\ref{sec:mutation_spectrum}),
and automatically identify  the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages})
with a high level of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of accurate joint estimation of
genetic inheritance with mutation and recombination will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

@@

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
% FIXME the development of thought is quite muddled here, we should rearrange
% in later revisions. We should discuss the expected lack of recombination
% and our likely false +s in one paragraph.
Compared to organisms like humans that recombine in every generation,
recombination is relatively rare in SARS-CoV-2, with recombination nodes accounting
for $<$0.3\% of the inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of strains. We expect ARGs to be particularly treelike
early in the pandemic, when co-infection was less likely and divergence between
lineages relatively low.

@@

The \texttt{sc2ts} backbone phylogenies for these Nextstrain subsamples contain small amounts of recombination,
with 7 recombination nodes in the Wide ARG backbone (8 for the Long ARG
backbone). However, the recombination events involve minor, local topological rearrangements
where recombination only occurs between close relatives.
(The majority of these recombinations are likely
false-positives, as discussed in Section~\ref{sec:false_positives}.)
